### 2021-02-25 rad-icfp-2021/background.tex
some more background

The problem with the previous approach shows up when one wants to compute the full gradient of a function $\RR^n\to\RR$, for a large $n$. 
Forward-mode only computes one directional derivative, for instance one partial derivative. This implies $n$ passes must be performed through the forward derivative to compute the whole gradient.
By using the symmetry in the chain rule, there is a way to compute the whole gradient faster, and this method is reverse-mode differentiation.
Suppose given a function $f=f_n\circ...\circ f_1 :\RR^n\to\RR$. 
The simplest way to see this mathematically, is first to say that forward mode essentially computes $(Jf)v=Jf_n(Jf_{n-1}(...(Jf_1v))...)$ for a direction $v\in\RR^n$. 
Reverse-mode, on the other hand, computes $(Jf)^Tv =Jf_1T^(Jf_{2}^T(...(Jf_n^Tv))...)$ for a vector $v\in\RR$. 
Because the compute flow of the function is reversed, the actual implementation of reverse-mode is way more tricky. 
%%TODO: from Matthijs' paper, need to rewrite.
Reverse-mode AD is only well-understood as a source-code transformation (also called define-then-run
style AD) on limited programming languages. Typically, its implementations
on more expressive languages that have features such as higher-order functions and conditionals
make use of define-by-run approaches. These approaches first build a computation graph during runtime, effectively evaluating the program until a straight-line
first-order program is left, and then they evaluate this new program \cite{carpenter2015stan,paszke2017automatic}. 
Such approaches have the severe downside that the differentiated code cannot benefit from existing optimizing compiler architectures. As such, these AD libraries
need to be implemented using carefully, manually optimized code, that for example does not contain any common subexpressions. This implementation process
is precarious and labour intensive. Further, some whole-program optimizations
that a compiler would detect go entirely unused in such systems.

@@

Following \cite{pearlmutter2008reverse}, there is an easy way to define an inefficient yet purely functional reverse-mode transformation for first-order programs.
\MH{TODO: finish}

@@

\begin{itemize}
    \item a
    \item  
\end{itemize}

>>>>

The problem with the previous approach shows up when one wants to compute the full gradient of a function $\RR^n\to\RR$, for a large $n$. 
Forward-mode only computes one directional derivative, for instance one partial derivative. This implies $n$ passes must be performed through the forward derivative to compute the whole gradient.
By using the symmetry in the chain rule, there is a way to compute the whole gradient faster, and this method is reverse-mode differentiation.
Suppose given a function $f=f_n\circ...\circ f_1 :\RR^n\to\RR$. 
The simplest way to see this mathematically, is first to say that forward mode essentially computes $(Jf)v=Jf_n(Jf_{n-1}(...(Jf_1v))...)$ for a direction $v\in\RR^n$. 
Reverse-mode, on the other hand, computes $(Jf)^Tv =J^Tf_1(J^Tf_{2}(...(J^Tf_nv))...)$ for a vector $v\in\RR$.
In particular, taking $v=1$ computes the gradient of $f$.

Because the compute flow of the function is reversed, the actual implementation of reverse-mode is way more tricky. 
%%TODO: from Matthijs' paper, need to rewrite.
Reverse-mode AD is only well-understood as a source-code transformation (also called define-then-run
style AD) on limited programming languages. Typically, its implementations
on more expressive languages that have features such as higher-order functions and conditionals
make use of define-by-run approaches. These approaches first build a computation graph during runtime, effectively evaluating the program until a straight-line
first-order program is left, and then they evaluate this new program \cite{carpenter2015stan,paszke2017automatic}. 
Such approaches have the severe downside that the differentiated code cannot benefit from existing optimizing compiler architectures. As such, these AD libraries
need to be implemented using carefully, manually optimized code, that for example does not contain any common subexpressions. This implementation process
is precarious and labour intensive. Further, some whole-program optimizations
that a compiler would detect go entirely unused in such systems.

@@

Following \cite{pearlmutter2008reverse}, there is a simple way to define an inefficient yet purely functional reverse-mode transformation for first-order programs.
We review a slight modification of the transformation we presented, which is also better explained on an example.

Let the term we want to differentiate be $x1:\RR,...,xn:\RR\vdash exp(cos(xi))$.
Following the chain rule, we need the jacobian matrices of $cos$ at $xi$ and of $exp$ at $cos(xi)$. 
Instead of considering these as operations from $\RR\to\RR$, we consider then as functions from the whole context. So $cos$ and $exp$ are seen as functions $\RR^n\to\RR$.
If we just do that though, we lose compositionality. 
So $cos$ will also return its context and is now seen as a function $\sem{cos}:\RR^n\to\RR^{n+1}$.
similarly for $exp$, but also taking the return value of $\sem{cos}$ as an extra argument, the one it will actually use and not simply return. 
We thus obtain $\sem{exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $J\sem{cos} \in Mat_{n,n+1}$, $J\sem{exp} \in Mat_{n+1,n+2}$ also compose nicely.
The same can be done for binary (or more) operators and let bindings (assuming no nested lets). 
This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ...\circ f_1$. 
If the original program was of type $\RR$, then the return value of the original program is the last compononent of $f$.
Following the mathematical presentation of reverse-mode above, the gradient of the the original program is then obtained as 
\[\nabla f= J^Tf(0,...,0,1)=J^Tf_1(J^Tf_{2}(...(J^Tf_m(0,...,0,1)))...)\]

To actually reverse the order of computation needed for this transpose of jacobians, we use a sort of simple continuation.
$f_i$ is turned into $\Dsynrevsymbol{f_i}:=<f_i, \lambda Y. Y\circ J^Tf_i>$ where $Y:\RR^{n+i-1}\to \RR^n$. 
We recover compositionality by noting that $<f_{i+1}(f_i), (\lambda Y. Y\circ J^Tf_{i+1}>)(\lambda Y. Y\circ J^Tf_i)>$ reduces to
$<f_{i+1}\circ f_i, \lambda Y. Y\circ J^Tf_i \circ J^Tf_{i+1}>$, and thus by induction we can obtain $<f, \lambda Y. YJ^Tf>$.
By applying the identity continuation $\RR^n\to\RR^n$ on the second component then the result to $(0,...,0,1)$, 
we have obtained a purely functional to compute $\nabla f$.

This method is quite inefficient though. First, we have to carry a continuation and $\beta$-reduce a lot of higher-order functions.
Second, each $J^Tf_{i+1}$ is a potentially huge matrix if $n$ or $m$ is big.

The imperative version of reverse mode for a binary operator $op(x,y)$ is something like $x'+= \partial_1op(x,y);y'+= \partial_2op(x,y)$. 
It is for these cases that mutation is usually key. 
If we see out term as a directed graph, reverse mode needs to backpropagate from the end of the graph to the starting nodes via every path.
It's hard to keep track of all these information in parallel in a functional way.
One of the key simple ideas that we used was to transform every operator into a unary one. 
This sssentially trivialises the compute flow to a line. 
Even if the starting program was a straight line program, 
having non unary operators was a source of inefficiency which justified mutation in the first place.
By returning every variable every time, the problem of using a variable several time does not need to be dealt with via mutation. 
This simple idea of transforming a program into essentially a straight line is what our new intermediate representation unary-form does.

@@

If we look at $J^Tf_{i+1}$, we notice that this function is almost the identity, except at the last row. 
Even on the last row, if the original term was a unary or binary operator like $cos, exp, +, *$, 
the row is zero except for at most 2 indices (1 for unary operators).
This means we can use a more compact representation $J^T\sem{op(xi,xj)}:=\lambda (y1,...,y(n+i)).(y1,...,y(n+i))+[i]\partial_1op(xi,xj)+[j]\partial_2op(xi,xj)$, 
where $[k]$ means that the element is added at the $k$-th index of the tuple.

We know in advance that such jacobian functions are going to be applied one to another, so we can use partial evaluation to $\beta$-reduce all of these $\lambda$s.
Because each function is almost the identity, we obtain a lot of substitutions of the form $[x/y]$ where both $x$ and $y$ are variables. 
This allows us to drastically reduce the size of $J^Tf$. In fact, for simple programs this is basically enough to get an efficient purely functional reverse derivative transformation.
We develop this idea further for a richer language.

### 2021-02-25 rad-icfp-2021/generalisation.tex
adding some figures
minor
outline paper in intro
small fixes
some more background

\section{Generalisations}

@@

\subsubsection{Higher-order functions}

>>>>

\section{Generalisations}
\label{sec:generalisation}

\MH{I think I only present two: conditionals, and general case for array. partial op can be dealt with before. HO function mentioned at the end}

Extended source language:

\input{figures/generalisation_source_grammar}

% \input{figures/generalisation_source_typesystem}

Note that we recover !zip A B! as !map2 (fun x,y -> (x,y)) A B)!. 
We can lift the restriction on the function of !foldl! by changing a part in $\invUNFSymbol$ of !foldl!.
\begin{center}
\begin{tabular}{r c l}
    !Y((x1,...,xm),...)! & -> &  !Y((x1,...,xm) + map2! \\
        && !(fun z1 z2 -> let E = map2 (fun x,y -> z2(x,y)) A e3 in! \\
        &&  !let F = zip B E in!\\
        && !foldl (fun (x,y), acc  -> x*acc+y) z1 F)! \\
        && $\nabla_{\Gamma}$!e2! $\nabla_{\Gamma}$!e1,...)!
\end{tabular}
\end{center}

Extended target language:

% \input{figures/generalisation_target_grammar}

% \input{figures/generalisation_target_typesystem}

@@

\subsection{Higher-order functions}
